---
title: 'Statistics 652: Statistical Learning and Prediction'
author: "Chidambaram Allada"
subtitle: Prediction of Departure Delay of Flights departing from NYC
output:
  word_document: default
  pdf_document: default
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)

```


  APPENDIX
  Software Details :
  R-Version- 3.6.1 \n
  LIst Of All Packages Required:\n
1)library(tidyverse) \n
2)library(nycflights13) \n
3)library(lubridate) \n
4)library(corrplot) \n
5)library(corrgram)\n
6)library(gam)\n
7)library(gbm)\n
8)library(xgboost)\n
9)library(tree)
10)library(randomForest)
11)library(dplyr)\n

Time to Knit : 
  



## Dataset
Dataset being analyzed is data from the nycflights13 package




```{r}
library(tidyverse)
library(nycflights13)
library(lubridate)
library(dplyr)
library(corrplot)
library(corrgram)
library(randomForest)
library(gam)
library(gbm)
library(xgboost)
library(tree)

#help(flights)
#help(weather)
#help(airports)
#help(planes)
fltrain <- read_csv("C:/Users/Chidu/Documents/fltrain.csv.gz")
fl_test <- read_csv("C:/Users/Chidu/Documents/fltest.csv.gz")
fltrain

```


```{r}
dim(fltrain)
```
Dimension of Training Dataset



## Handling of Missing data




```{r}
fl <- fltrain
for(i in 1:ncol(fl)) {
  if(typeof(fl[[i]]) == "character") {
    fl[[i]] <- factor(fl[[i]])
  }
}

```


counting the missing values in each variable.

```{r}
missing_value_features <- function(x) { sum(is.na(x)) }
sapply(fl,missing_value_features)
```

Discarding the variables that have more than 5% missing values which is around 10000.

```{r}
fl <- fl%>% select(-year.y,-type,-manufacturer,-model,-engines,-seats, -speed, -engine,-wind_gust,-pressure)
summary(fl)
dim(fl)

```

na.omit used to omit the missing values 

```{r}
fl <- na.omit(fl)
dim(fl)
summary(fl)

```

## Summaries of the response variable `dep_delay`

The departure delays variable is highly right-skewed that is showed.

```{r}
range(fl$dep_delay)
fivenum(fl$dep_delay)
quantile(fl$dep_delay,probs = c(0,0.01,0.05,0.1,0.25,.5,.75,.90,.95,.99,1))
mean(fl$dep_delay >= 60) # about 15,000 or 8% of flights
```

Top 10 delays.

```{r}
fl%>% arrange(desc(dep_delay)) %>% head(10) 
```


Summaries of departure delay by NYC airport:
```{r}
Q3 <- function(x) { quantile(x,probs=.75) }
fl %>% group_by(origin) %>% 
  summarize(n=n(),med_d = median(dep_delay),Q3_d = Q3(dep_delay), max_d = max(dep_delay)) %>% 
  arrange(desc(Q3_d)) %>% head(10) 
```

Summaries of departure delay by airline (carrier).

```{r}
fl %>% group_by(carrier) %>% 
  summarize(n=n(),med_d = median(dep_delay),Q3_d = Q3(dep_delay), max_d = max(dep_delay)) %>% 
  arrange(desc(Q3_d)) %>% head(10) 
fl %>% group_by(origin,carrier) %>% 
  summarize(n=n(),med_d = median(dep_delay),Q3_d = Q3(dep_delay), max_d = max(dep_delay)) %>% 
  arrange(desc(Q3_d)) %>% head(10) 
fl %>% group_by(dest,carrier) %>% 
  summarize(n=n(),med_d = median(dep_delay),Q3_d = Q3(dep_delay), max_d = max(dep_delay)) %>% 
  arrange(desc(Q3_d)) %>% head(10) 
```

Summaries of departure delay by date:

```{r}
fl %>% group_by(month,day) %>% 
  summarize(n=n(),med_d = mean(dep_delay),max_d = max(dep_delay)) %>% 
  arrange(desc(med_d)) %>% head(10) # what happened on march 8?
```


Summaries of departure delay by precipitation:
```{r}
fl %>% mutate(haveprecip = factor(precip>0)) %>% group_by(haveprecip) %>% 
  summarize(n=n(),med_d = median(dep_delay),Q3_d = Q3(dep_delay), max_d = max(dep_delay)) %>% 
  arrange(desc(med_d)) %>% head(10) 
```



```{r}

den <- nrow(fl)+1
fl <- fl %>% mutate(dep_delay = rank(dep_delay)/den)
ggplot(fl,aes(x=dep_delay)) + geom_histogram(binwidth=.01)

```

# Data Wrangling Part




```{r}

fl <- fl %>% 
  mutate(dep_date = make_date(year.x,month,day)) %>% 
  select(-year.x,-month,-day,-dep_time,-arr_time,-arr_delay,
         -sched_arr_time,-tailnum,-flight,-name,-air_time,
         -hour,-minute,-time_hour,-tz,-dst,-dest) %>%
  mutate(precip = as.numeric(precip>0))
corrplot(corrgram(fl),type = "upper", order = "hclust",tl.col = "black", tl.srt = 45)
```


## Relationship between `dep_delay` and quantitative predictors



```{r,cache=TRUE, eval= FALSE}
ggplot(fl,aes(x=dep_date,y=dep_delay)) + geom_point(alpha=.01) + geom_smooth()

ggplot(fl,aes(x=sched_dep_time,y=dep_delay)) + geom_point(alpha=0.01) + geom_smooth()

ggplot(fl,aes(x=distance,y=dep_delay)) + geom_point(alpha=0.01) + geom_smooth()
ggplot(fl,aes(x=log(distance),y=dep_delay)) + geom_point(alpha=0.01) + geom_smooth()


ggplot(fl,aes(x=temp,y=dep_delay)) + geom_point(alpha=0.01) + geom_smooth()

ggplot(fl,aes(x=dewp,y=dep_delay)) + geom_point(alpha=0.01) + geom_smooth()


```



```{r,cache=TRUE}

fl <- mutate(fl,logdistance = log(distance)) %>% select(-distance)
fl <- mutate(fl,logalt = log(alt)) %>% select(-alt)

```


## Splitting the Training set into train and validation set for tuning of the parameters



```{r}
set.seed(123)
tzone1 <- as.factor(fl$tzone)
fl <- mutate(fl,tzone=tzone1)
tr_size <- ceiling(2*nrow(fl)/3)
train <- sample(1:nrow(fl),size=tr_size)
fl_tr <- fl[train,]
fl_te <- fl[-train,]

var_dd <- var(fl_te$dep_delay)
var_dd
```

## Transforming the Test data set as per the trained model 

```{r}
set.seed(9)

for(j in 1:ncol(fl_test)) {
  if(typeof(fl_test[[j]]) == "character") {
    fl_test[[j]] <- factor(fl_test[[j]])
  }
}

fl_test <- fl_test%>% select(-year.y,-type,-manufacturer,-model,-engines,-seats, -speed, -engine,-wind_gust,-pressure)
fl_test <- na.omit(fl_test)
dim(fl_test)
den_test <- nrow(fl_test)+1
fl_test <- fl_test %>% mutate(dep_delay = rank(dep_delay)/den_test)

fl_test <- fl_test %>% 
  mutate(dep_date = make_date(year.x,month,day)) %>% 
  select(-year.x,-month,-day,-dep_time,-arr_time,-arr_delay,
         -sched_arr_time,-tailnum,-flight,-name,-air_time,
         -hour,-minute,-time_hour,-tz,-dst,-dest) %>%
  mutate(precip = as.numeric(precip>0))

fl_test <- mutate(fl_test,logdistance = log(distance)) %>% select(-distance)
fl_test <- mutate(fl_test,logalt = log(alt)) %>% select(-alt)

tzone2 <- as.factor(fl_test$tzone)
carrier_test <-as.factor(fl_test$carrier)
origin_test <- as.factor(fl_test$origin)
fl_test <- mutate(fl_test,carrier=carrier_test,origin=origin_test,tzone=tzone2)

dep_date_numeric_t <- as.numeric(fl_test$dep_date)
dep_date_numeric_t <- dep_date_numeric_t - mean(dep_date_numeric_t)
fl_test <- mutate(fl_test,dep_date = dep_date_numeric_t)



```




```{r,cache=TRUE, eval=FALSE}

set.seed(67)
form <- formula(dep_delay ~ s(dep_date) + s(sched_dep_time) + carrier + origin + tzone + s(logdistance) +
                  s(temp) + s(dewp) + s(humid) + s(wind_dir) + s(wind_speed) + precip + s(visib))
gam_fit <- gam(form, data=fl_tr,family=gaussian) 
summary(gam_fit)
plot(gam_fit,se=TRUE)
gam_pred <- predict(gam_fit,newdata=fl_te)
mse_gam <- mean((fl_te$dep_delay-gam_pred)^2)

mse_gam
abs(mse_gam - var_dd)/var_dd
```




```{r,cache=TRUE}

dep_date_numeric <- as.numeric(fl_tr$dep_date)
dep_date_numeric <- dep_date_numeric - mean(dep_date_numeric)
fl_tr_tem <- mutate(fl_tr,dep_date = dep_date_numeric)

dep_date_numeric <- as.numeric(fl_te$dep_date)
dep_date_numeric <- dep_date_numeric - mean(dep_date_numeric)
fl_te_tem <- mutate(fl_te,dep_date = dep_date_numeric)


```



```{r,cache=TRUE,eval=FALSE}

set.seed(142)

gbm_fit <-gbm(dep_delay ~ .,data=fl_tr_tem,distribution="gaussian",
              n.trees = 2000, shrinkage = 0.2, interaction.depth = 3)
summary(gbm_fit)

gbm_pred <- predict(gbm_fit,newdata=fl_te_tem,n.trees = 2000)
gbm_pred_test <- predict(gbm_fit,newdata=fl_test,n.trees = 2000)
mse_gbm <- mean((fl_te$dep_delay-gbm_pred)^2)
mse_gbm_test <- mean((fl_test$dep_delay-gbm_pred_test)^2)
mse_gbm
mse_gbm_test

abs(mse_gbm - var_dd)/var_dd

```



Decision trees for regression
```{r}

set.seed(123)

tree.model <- tree(dep_delay~., fl_tr_tem)
summary(tree.model)
plot(tree.model)
text(tree.model,pretty=0)
cv.model = cv.tree(tree.model)
plot(cv.model$size,cv.model$dev, type = "b")
```



```{r}
prune.model= prune.tree(tree.model, best=4)
yhat=predict(prune.model, newdata=fl_te_tem)
yhat_test = predict(prune.model, newdata=fl_test)
mean((yhat-fl_te_tem$dep_delay)^2)
mean((yhat_test-fl_test$dep_delay)^2)
```

Using Random Forest method for regression


```{r, eval=FALSE}

set.seed(432)
fl_rf <- fl_tr_tem[1:35000,]
bag.flights <- randomForest(fl_rf$dep_delay~., data=fl_rf,ntree=100, importance=TRUE,na.action=na.omit)
bag.flights
yhat.bag <- predict(bag.flights,newdata= fl_te_tem)
yhat.bag_test <- predict(bag.flights, newdata=fl_test)
mean((yhat.bag-fl_te_tem$dep_delay)^2)
mean((yhat.bag_test -fl_test$dep_delay)^2)

```
```{r}

set.seed(134)
data.xg <- xgb.DMatrix(data=data.matrix(fl_tr_tem[-2]), label=fl_tr_tem$dep_delay)
bst_dmatrix <- xgboost(data=data.xg, max.depth=10,eta = 0.3, nrounds=100, eval_metric ="rmse")
bst_dmatrix
data.val.xg <- xgb.DMatrix(data=data.matrix(fl_te_tem[-2]), label=fl_te_tem$dep_delay)
data.test.xg <- xgb.DMatrix(data=data.matrix(fl_test[-2]), label=fl_test$dep_delay)
pred <- predict(bst_dmatrix,data.val.xg)
pred_test <- predict(bst_dmatrix,data.test.xg)
xgb_val_mse <- mean((pred-fl_te_tem$dep_delay)^2)
xgb_val_mse
xgb_te_mse <- mean((pred_test-fl_test$dep_delay)^2)
xgb_te_mse
```
